Sampling Strategies in Decisions from Experience

### 2022-04-04 manuscript.Rmd
Review Figure and title page

---
title             : "Sampling Strategies in Decisions from Experience: Property Representation and Probability Weighting Approximation"
shorttitle        : "Sampling strategies"

author: 
  - name          : "Linus Hof"
    affiliation   : "1"
    corresponding : yes    
    address       : "Department of Psychology, Heidelberg University, Hauptstraße 47-51, 69117 Heidelberg, Germany"
    email         : "linushof@posteo.de"

affiliation:
  - id            : "1"
    institution   : "Heidelberg University"

note: | 
  **Unsubmitted draft created from the commit with the hash**: ``r repro::current_hash()``

authornote: |
 This is a dynamic document which can be reproduced from the accompanying GitHub repository: https://github.com/linushof/sampling-in-dfe.

 The current version of this manuscript is prepared for submission as a master's thesis.
 Supervisors: Thorsten Pachur & Veronika Zilker.

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, #omit messages
                      fig.align = "l", #align figures to left margin
                      fig.pos = "t" # display figures at the top of a page
                      )
```

@@

```{r weighting-function, fig.cap="(ref:weighting-function)", fig.height=7, fig.width=10}
#compute images of weighting function (wf)
wf <- tibble(p = seq(0, 1, .01)) %>%  #cumulative probabilities
  expand_grid(gamma = seq(.1, 2, .1), #gamma values
              delta = c(.1, .5, 1, 2, 5, 10)) %>% #delta values
  mutate(wp = (delta*(p^gamma))/((delta*p^gamma)+(1-p)^gamma)) #images of wf

@@

#plot shapes of weighting function
wf %>% 
  ggplot(aes(p, wp, group = gamma)) +
  facet_wrap(~delta, labeller = as_labeller(label_delta, default = label_parsed)) + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x = expression(p),
       y = expression(w(p)), 
       color = expression(gamma)) + 
  theme_apa() + 
  geom_line(aes(color = gamma)) +
  scale_color_viridis(option = "viridis") + 
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed")
```

@@

```{r data, include=FALSE}
#load choice data
cols_choices <- list(.default = col_double(),
                     boundary = col_factor(),
                     gamble = col_factor(),
                     rare = col_factor(),
                     agent = col_factor(),
                     choice = col_factor())
choices <- read_csv("data/choices/choices.csv", col_types = cols_choices)

#load CPT estimates
cols_cpt <- list(.default = col_double(),
                 boundary = col_factor(),
                 a = col_factor(),
                 parameter = col_factor())
cpt <- read_csv("data/estimates/estimates_cpt.csv", col_types = cols_cpt)
```

@@

```{r}
choices %>%
  group_by(s, a) %>% 
  summarize(n_med = median(n_sample)) %>% 
  ggplot(aes(x = s, y = n_med, group = as.factor(a), color = as.factor(a))) + 
   labs(x = expression(psi),
       y = "Sampled Outcomes",
       color = expression(theta)) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 50)) +
  geom_point() +
  geom_line() + 
  scale_color_viridis_d(option = "viridis") +
  theme_apa()
```

<!-- Plots choice rates -->

(ref:choice-rates) Effect of Switching Probability on Rates of False Risky and False Safe Choices for 2-Outcome Prospects with High Rank Outcomes of Different Probability

```{r choice-rates, fig.cap="(ref:choice-rates)", fig.height=7, fig.width=10}
#prepare data

##determine normative choice according to latent and sampled EV
fr_rates <- choices %>%
  mutate(norm = case_when(ev_ratio > 1 ~ "A", ev_ratio < 1 ~ "B"), #normative choice latent EV
         ev_ratio_exp = round(a_ev_exp/b_ev_exp, 2),
         norm_exp = case_when(ev_ratio_exp > 1 ~ "A", ev_ratio_exp < 1 ~ "B")) #normative choice sampled EV

##compute false response rates

###latent EV
fr_rates_l <- fr_rates %>%
  filter(!is.na(norm)) %>% #exclude trials with equal EV
  group_by(s, boundary, a, rare, norm, choice) %>% #separate by model parameter and type of rare event
  summarise(n = n()) %>%
  mutate(rate = round(n/sum(n), 2), #compute response rates
         type = case_when(norm == "A" & choice == "B" ~ "Safe", #determine false response type
                          norm == "B" & choice == "A" ~ "Risky")) %>%
  ungroup() %>%
  filter(!is.na(type)) %>% #remove correct responses
  select(-c(norm, choice, n)) %>%
  mutate(norm_ev = "Latent") #assign facet label

###sampled EV
fr_rates_s <- fr_rates %>%
  filter(!is.na(norm_exp)) %>%
  group_by(s, boundary, a, rare, norm_exp, choice) %>%
  summarise(n = n()) %>%
  mutate(rate = round(n/sum(n), 2),
         type = case_when(norm_exp == "A" & choice == "B" ~ "Safe",
                          norm_exp == "B" & choice == "A" ~ "Risky")) %>%
  ungroup() %>%
  filter(!is.na(type)) %>%
  select(-c(norm_exp, choice, n)) %>%
  mutate(norm_ev = "Sample")

### merge latent and sampled data
fr_rates <- bind_rows(fr_rates_l, fr_rates_s) %>%
  mutate(rare = case_when(rare == "none" ~ "\\in \\[.2,.8\\]", #change facet labels for rare events
                          rare == "attractive" ~ "\\in (0,.2)",
                          rare == "unattractive" ~ "\\in (.8,1)"))

@@

#plot
fr_rates %>%
  ggplot(aes(s, rate)) +
  facet_grid(rare~a, labeller = labeller(rare = as_labeller(label_rare, default = label_parsed), 
                                         a = as_labeller(label_theta, default = label_parsed))) +
  labs(x = "Switching Probability",
       y = "False Response Rate",
       color = "Normative\nExpected\nValue",
       shape = "False\nResponse\nType") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  geom_point(aes(color = norm_ev, shape = type)) +
  scale_shape_manual(values = c(4, 16)) +
  scale_color_viridis_d(option = "viridis", begin = .1, end = .7) +
  theme_apa()
```

<!-- Plots CPT -->

(ref:weighting-parameters) Parameter Estimates of the Weighting Function of @goldsteinExpressionTheoryPreference1987 for Different Simulation Parameters

```{r weighting-parameters, fig.cap="(ref:weighting-parameters)", fig.height=7, fig.width=10}
#prepare data
cpt <- cpt %>% mutate(boundary = if_else(boundary == "absolute", "Absolute", "Relative"))

@@

##gamma
gamma <- cpt %>%
  filter(parameter == "gamma") %>%
  filter(a != 1, boundary != "relative") %>% 
  ggplot(aes(s, mean, color = s, shape = boundary)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) +
  labs(x = element_blank(), #omit axis title
       y = expression(gamma),
       color = "Switching\nProbability",
       shape = "Boundary") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,.5)) +
  geom_point() +
  geom_pointrange(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  scale_shape_manual(values = c(4, 16)) +
  scale_color_viridis(option = "viridis") +
  theme_apa()

##delta
delta <- cpt %>%
  filter(parameter == "delta") %>% 
  filter(a != 1, boundary != "relative") %>% 
  ggplot(aes(s, mean, color = s, shape = boundary)) +
  facet_wrap(~a, nrow = 1) + 
  labs(x = "Switching Probability",
       y = expression(delta),
       color = "Switching\nProbability",
       shape = "Boundary") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,.5)) +
  geom_point() +
  geom_pointrange(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  scale_color_viridis(option = "viridis") +
  scale_shape_manual(values = c(4, 16)) +
  theme_apa() + 
  theme(strip.text.x = element_blank()) #omit facet labels

##merge figures
ggarrange(gamma, delta, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right") 
```

(ref:value-parameters) Parameter Estimates of the Value Function of @tverskyAdvancesProspectTheory1992 and of the Logit Choice Rule for Different Simulation Parameters

```{r value-parameters, fig.cap="(ref:value-parameters)", fig.height=7, fig.width=10}
#plot estimates

##alpha
alpha <- cpt %>%
  filter(parameter == "alpha") %>% 
  filter(a != 1) %>% 
  ggplot(aes(s, mean, color = s, shape = boundary)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) +
  labs(x = element_blank(),
       y = expression(alpha),
       color = "Switching\nProbability",
       shape = "Boundary") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,.5)) +
  geom_point() +
  geom_pointrange(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  scale_color_viridis(option = "viridis") +
  scale_shape_manual(values = c(4, 16)) +
  theme_apa()

##rho
rho <- cpt %>%
  filter(parameter == "rho") %>% 
  filter(a != 1) %>% 
  ggplot(aes(s, mean, color = s, shape = boundary)) +
  facet_wrap(~a, nrow = 1) + 
  labs(x = "Switching Probability",
       y = expression(rho),
       color = "Switching\nProbability",
       shape = "Boundary") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1,.5)) +
  geom_point() +
  geom_pointrange(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  scale_color_viridis(option = "viridis") +
  scale_shape_manual(values = c(4, 16)) +
  theme_apa() + 
  theme(strip.text.x = element_blank())

##merge figure
ggarrange(alpha, rho, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right") 
```

@@

```{r cpt-graphs, fig.cap="(ref:cpt-graphs)", fig.height=7, fig.width=10}
#prepare data

>>>>

---
title             : |
  Sampling Strategies in Decisions from Experience: Property Representation and Probability Weighting Approximation

shorttitle        : "Sampling strategies"

author: 
  - name          : "Linus Hof"
    corresponding : yes    
    email         : "linushof@posteo.de"

note: |
    Unsubmitted draft created from the commit `r repro::current_hash()`

authornote: |
 The article is prepared for submission as a master's thesis in fulfillment of the requirements for the degree Master of Science (M.Sc.) at

 Heidelberg University

 Faculty of Behavioural and Cultural Studies

 Institute of Psychology

  \noindent
  under the supervision of

  \indent
  1. Prof. Dr. Thorsten Pachur, Technical University of Munich

  \indent
  2. Dr. Veronika Zilker, Max Planck Institute for Human Development

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, #omit messages
                      fig.align = "l", #align figures to left margin
                      fig.pos = "t", # display figures at the top of a page
                      fig.height=7, 
                      fig.width=10)
```

@@

```{r weighting-function, fig.cap="(ref:weighting-function)"}
#compute images of weighting function (wf)
wf <- tibble(p = seq(0, 1, .01)) %>%  #cumulative probabilities
  expand_grid(gamma = seq(.1, 2, .1), #gamma values
              delta = c(.1, .5, 1, 2, 5, 10)) %>% #delta values
  mutate(wp = (delta*(p^gamma))/((delta*p^gamma)+(1-p)^gamma)) #images of wf

@@

#plot shapes of weighting function
wf %>% 
  ggplot(aes(p, wp, group = gamma)) +
  facet_wrap(~delta, labeller = as_labeller(label_delta, default = label_parsed)) + 
  scale_x_continuous(breaks = seq(0, 1, .5)) +
  scale_y_continuous(breaks = seq(0, 1, .5)) + 
  labs(x = expression(p),
       y = expression(w(p)), 
       color = expression(gamma)) + 
  theme_apa() + 
  geom_line(aes(color = gamma), size = .5) +
  scale_color_viridis(option = "viridis") + 
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed")
```

@@

```{r data, include=FALSE}
#load choice data
cols_choices <- list(.default = col_double(),
                     boundary = col_factor(),
                     gamble = col_factor(),
                     rare = col_factor(),
                     agent = col_factor(),
                     choice = col_factor())
choices <- read_csv("data/choices/choices.csv", col_types = cols_choices)
choices <- choices %>% 
  filter(boundary == "absolute")

#load CPT estimates
cols_cpt <- list(.default = col_double(),
                 boundary = col_factor(),
                 a = col_factor(),
                 parameter = col_factor())
cpt <- read_csv("data/estimates/estimates_cpt.csv", col_types = cols_cpt)
cpt <- cpt %>% 
  filter(boundary == "absolute")
```

@@

We start by inspecting how the number of outcomes sampled over the course of a sampling and accumulation process changes with varying combinations of switching probabilities and thresholds.
Figure\ \@ref(fig:chain-length) shows the respective median for each parameter combination.
The model predicts an inverse relationship between the switching probability and the overall number of sampled outcomes, since the stochastic processes tend to terminate faster, the higher the switching probabilities.
In other words, the higher the switching probability, the smaller the samples on which each mean comparison is based, resulting in a decreasing overall number of sampled outcomes.
Since the relative change in the switching probability decreases as $\psi$ approaches certainty, the relative change in the number of sampled outcomes decreases respectively.
Furthermore, because it determines the number of mean comparisons---and therefore stochastic processes---that must be carried out, any increase in the threshold leads to an increase in the number of sampled outcomes.

(ref:chain-length) Lengths of Information-Processing Chains

```{r chain-length, fig.cap="(ref:chain-length)"}
choices %>%
  group_by(s, a) %>% 
  summarize(n_med = median(n_sample)) %>% 
  ggplot(aes(x = s, y = n_med, group = as.factor(a), color = as.factor(a))) + 
   labs(x = expression(psi),
       y = "Median of Overall Sampled Outcomes",
       color = expression(theta)) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_y_continuous(limits = c(0, 120), breaks = seq(0, 150, 50)) +
  geom_point() +
  geom_line() + 
  scale_color_viridis_d(option = "viridis") +
  theme_apa()
```

### Final Choice

Figure\ \@ref(fig:choice-rates) displays the rates of choices that did not maximize the EV for each combination of the switching probabilities and thresholds.
The top panel of the figure shows the rates for choice problems where the low-rank outcome of the risky prospect has a low probability.
Respectively, for high switching probabilities and therefore short stochastic processes, the low-rank outcome is expected to occur less often over the course of a stochastic process than would be objectively warranted, leading to an inflation of the respective mean of the risky prospect.
Accordingly, if the safe prospect possesses the higher EV, the number of false responses increases with the switching probability.
However, if the risky prospect has the higher EV, the mean inflation causes the choices to be generally in line with EV maximization.
Note from the middle panel that this pattern reverses, if the high-rank outcome has a small probability.

(ref:choice-rates) Rates of False Risky and False Safe Choices

```{r choice-rates, fig.cap="(ref:choice-rates)"}
#prepare data

##determine normative choice according to latent and sampled EV
fr_rates <- choices %>%
  mutate(norm = case_when(ev_ratio > 1 ~ "A", ev_ratio < 1 ~ "B")) %>% # determine normative choice
  filter(!is.na(norm)) %>% #exclude trials with equal EV
  group_by(s, boundary, a, rare, norm, choice) %>% #separate by model parameter and type of rare event
  summarise(n = n()) %>%
  mutate(rate = round(n/sum(n), 2), #compute response rates
         type = case_when(norm == "A" & choice == "B" ~ "Safe", #determine false response type
                          norm == "B" & choice == "A" ~ "Risky")) %>%
  ungroup() %>%
  filter(!is.na(type)) %>% #remove correct responses
  select(-c(norm, choice, n)) %>% 
  mutate(rare = case_when(rare == "none" ~ "\\in \\[.2,.8\\]", #change facet labels for rare events
                          rare == "attractive" ~ "\\in (0,.2)",
                          rare == "unattractive" ~ "\\in (.8,1)"))

@@

#plot
fr_rates %>%
  ggplot(aes(s, rate, group = type)) +
  facet_grid(rare~a, labeller = labeller(rare = as_labeller(label_rare, default = label_parsed), 
                                         a = as_labeller(label_theta, default = label_parsed))) +
  labs(x = expression(psi),
       y = "False Response Rate",
       color = expression(psi),
       shape = "False\nResponse") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, .5)) +
  geom_line(aes(color = s, group = type)) + 
  geom_point(aes(color = s, shape = type), size = 2) +
  scale_shape_manual(values = c(4, 16)) +
  scale_color_viridis(option = "viridis") +
  theme_apa()
```

### Cumulative Prospect Theory

(ref:weighting-parameters) Parameter Estimates and Graphical Shape of the Weighting Function

```{r weighting-parameters, fig.cap="(ref:weighting-parameters)"}

@@

##gamma
gamma <- cpt %>%
  filter(parameter == "gamma") %>%
  ggplot(aes(s, mean, color = s)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) +
  scale_y_continuous(limits = c(0,2), breaks = seq(0,2.0,1.0))+
  labs(x = element_blank(), #omit axis title
       y = expression(gamma),
       color = expression(psi)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  scale_color_viridis(option = "viridis") +
  theme_apa() +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())

##delta
delta <- cpt %>%
  filter(parameter == "delta") %>% 
  ggplot(aes(s, mean, color = s)) +
  facet_wrap(~a, nrow = 1) + 
  scale_y_continuous(limits = c(0,10), breaks = seq(0,10,5)) +
  scale_x_continuous(limits = c(0, 1.1), breaks = seq(0,1,.5)) +
  labs(x = expression(psi),
       y = expression(delta),
       color = expression(psi)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point() +
  scale_color_viridis(option = "viridis") +
  theme_apa() + 
  theme(strip.text.x = element_blank()) #omit facet labels

wf <- cpt %>%
  select(s, boundary, a, parameter, mean) %>%
  pivot_wider(names_from = parameter, values_from = mean) %>%
  select(-c(alpha, rho)) %>%
  expand_grid(p = seq(0, 1, .05)) %>%
  mutate(w = round(  (delta * p^gamma)/ ((delta * p^gamma)+(1-p)^gamma), 2))

##weighting function
weight <- wf %>% #graphs for absolute boundary
  ggplot(aes(p, w, group = s, color = s)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) + 
  labs(x = "Experienced Probability",
       y = expression(pi),
       color = expression(psi)) +
  scale_x_continuous(breaks = seq(0, 1, .5)) +
  scale_y_continuous(breaks = seq(0, 1, .5)) +
  geom_line(size = .5) +
  scale_color_viridis(option = "viridis") +
  theme_apa() + 
  theme(strip.text.x = element_blank()) #omit facet labels

##merge figures
ggarrange(gamma, delta, weight, ncol = 1, nrow = 3, common.legend = TRUE, legend = "right")
```

(ref:value-parameters) Parameter Estimates and Graphical Shape of the Value Function

```{r value-parameters, fig.cap="(ref:value-parameters)"}
#plot estimates

##alpha
alpha <- cpt %>%
  filter(parameter == "alpha") %>% 
  ggplot(aes(s, mean, color = s)) +
  facet_wrap(~a, nrow = 1, labeller = labeller(a = as_labeller(label_theta, default = label_parsed))) +
  scale_x_continuous(limits = c(0, 1.1), breaks = seq(0,1,.5)) +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1, .5)) + 
  labs(x = expression(psi),
       y = expression(alpha),
       color = expression(psi)) +
  geom_point() +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  scale_color_viridis(option = "viridis") +
  theme_apa()

value <- vf %>% 
  ggplot(aes(x, v, group = s, color = s)) +
  facet_wrap(~a, nrow = 1)+ 
  labs(x = "Objective Outcome",
       y = "Subjective Value",
       color=expression(psi)) +
  scale_x_continuous(limits = c(0,20), breaks = seq(0, 20, 10)) +
  scale_y_continuous(limits = c(0,20), breaks = seq(0, 20, 10)) +
  geom_line(size = .5) +
  scale_color_viridis(option = "viridis") +
  theme_apa() + 
  theme(strip.text.x = element_blank())

##rho
rho <- cpt %>%
  filter(parameter == "rho") %>% 
  ggplot(aes(s, mean, color = s, shape = boundary)) +
  facet_wrap(~a, nrow = 1) + 
  labs(x = "Switching Probability",
       y = expression(rho),
       color = "Switching\nProbability",
       shape = "Boundary") +
  scale_x_continuous(limits = c(0, 1.1), breaks = seq(0, 1,.5)) +
  geom_point() +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  scale_color_viridis(option = "viridis") +
  scale_shape_manual(values = c(4, 16)) +
  theme_apa() + 
  theme(strip.text.x = element_blank())

##merge figure
ggarrange(alpha, value, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right") 
```

@@

```{r cpt-graphs, fig.cap="(ref:cpt-graphs)", eval = FALSE}
#prepare data
